Ground state energy of quantum dots using the coupled cluster method

### 2018-06-14 project_2/article/article.tex
Add reviewed project

    Regardless of how sophisticated the method used, the many-body problem
    remains quite intractable. The true wavefunction of any system will be a
    linear combination of all possible slater-type orbitals one can model. This
    means that in order to represent a system correctly, one would need a basis
    set of infinite size. This is an obvious problem when we solve problems
    numerically, because we obviously need to limit the size of the basis we use
    because we have limitations in memeory and computating power. Luckily, great
    minds have bestowed a wide selection of very god numerical methods upon the
    world, which the coupled cluster is a part of.

    First, we provide a thorough theoretical outline of how we model a quantum
    dot, the mathematical formalism behind many-body problems and outline how
    the computational methods we employ both turn into iterative schemes that
    are easy to represent numerically. Second, we provide a brief overview of
    the program we have constructed which is in the form an installable python
    package.  Third, we have employed the methods on a large parameter space
    presenting results for systems consisting of $N= \{2, 6, 12, 20\}$
    particles. Fourth, we discuss the most interesting aspects of the results
    and try to shed some light over the shortcomings as well as the beneficial
    features of the different metods. Lastly, we end on some summary remarks.

@@

        The theoretical approximation of a quantum dot is impossible to make
        perfectly accurate, yet simple. There are some important effects that
        should be included such the potential arising as a result of the
        confinement, and the interaction between particles. We therefore model
        our system with an electron-electron interaction, given by the two-body
        operator in \autoref{eq:two_body_operator}, and with we approximate the
        confining potential by the harmonic oscillator potential, given by the
        second term in the one-body Hamiltonian in
        \autoref{eq:one_body_hamiltonian}. The last term in the one-body
        Hamiltonian is the kinetic energy of particle in question, essential in
        any system.

@@

    \subsection{Implementation details}
        Both in \texttt{CoupledClusterDoublesSparse} and
        \texttt{CoupledClusterDoublesOptimized} we make use of the
        intermediates in equations \ref{eq:intermediate1} to
        \ref{eq:intermediate4}.  In the optimized class special care was taken
        to ensure that every contraction was computed in an optimal way, with
        the minimization of memory use and CPU time as the overall goal.
        Moreover, several methods in NumPy for doing linear algebra was tested,
        including but not limited to \texttt{numpy.einsum},
        \texttt{numpy.tensordot}, and reshaping datastructures to
        two dimensions and using \texttt{numpy.matmul}.
        Additionally, we tried to make use of the python package
        \texttt{dask} which allows for arrays that are larger than
        the computer's memory buffer as well as automated concurrency. We found
        that the dask arrays was a bad fit for our code base, but we ended up
        using \texttt{numba} for paralellization of some
        iterative loops as well as just-in-time compilation of some functions.

    \subsection{Scaling of the coupled cluster doubles method}
        When talking about the scaling of the coupled cluster method there are
        two aspects to consider.
        \begin{enumerate}
            \item The number of FLOPS to compute $g(f, u, t)$.
            \item The number of bytes needed to store the arrays of matrix
                elements.
        \end{enumerate}
        For the CCD method the naïve solution using the form of $g(f, u, t)$
        without the intermediates results in $\mathcal{O}(M^4N^4)$ FLOPS each
        iteration. Using the form of the intermediates we have set up this
        reduces to $\mathcal{O}(M^4N^2)$, i.e., two order of magnitude less. The
        effect of this is \emph{dramatic} and will be shown in the results
        section in due time.

@@

    \subsection{Validity of results}
        The analytical computed energy for $\omega=1.0$ a.u.
        and $N=2$ is $3$\cite{taut1994two}. From our tables we see
        that RHF is least successful in reaching this value, CCD in
        HF basis is best, while CCD with HO basis is generally somewhere
        between the two. We see that we get even closer to the analytical
        ``truth'' by increasing the number of shells.

        The rest of the results have been compared with the master thesis of M.
        P.  Lohne\cite{lohne2010coupled} for up to $R = 10$ shells. Lohne gets
        two sets of energies, one set from RHF and one set from CCSD code with
        harmonic oscillator basis functions. Our CCD code with Hartree-Fock
        basis will for some configurations\footnote{By configurations we mean
        frequency $\omega$ and number of particles $N$.} beat CCSD with plain
        harmonic oscillator basis. For $R = 12$ shells we can compare with the
        results from M. P. Lohne et al.\cite{lohne2011ab}, but in this article
        an \emph{effective interaction} for the Hamiltonian with a CCSD code
        using the HF basis has been used. These results are therefore
        significantly better than ours, but we can get a ``ball-park'' idea to
        benchmark against.

    \subsection{Running time}
        Looking at the running times shown in \autoref{fig:running_time_2} and
        \autoref{fig:running_time_6} we can see how the naïve implementation
        blows up very fast. As both the sparse and the optimized version uses
        intermediate calculations these running times scales on the order of
        $\mathcal{O}(l^6)$ whereas the naïve implementation goes as
        $\mathcal{O}(l^8)$. Another important factor is that the naïve
        implementation uses \texttt{np.einsum} to calculate the tensor
        contractions whereas the sparse and optimized version uses
        \texttt{sparse.tensordot} and \texttt{np.matmul} respectively. The two
        latter function calls uses BLAS's implementation of the matrix dot which
        is way faster than Numpys \texttt{np.einsum}.

    \subsection{Convergence trouble}
        For a large number of particles and a low frequency the CCD-method get
        trouble with convergence. This happens as the confining potential $v
        \propto \omega^2$ will not be able to confine the particles when the
        interaction gets too strong. The CCD-method will in particular get a
        hard time keep the particles together as the excitation operator $t$
        only excites pairs of particles leading to a too strong increase in the
        interaction when the particles increase their energy level. This effect
        can potentially be somewhat alleviated by including the singles
        excitation, e.g., using the CCSD-method.

        We can see this behaviour in our results. For $N = 2$ we get convergence
        for all methods for $\omega = 0.1$ shown in \autoref{tab:N2}. By
        increasing to $N = 6$ particles we immediately see how CCD with the HO
        basis will get in trouble for small values of $\omega$. Even by increasing
        the strength of the confinment this method does converge.

        For $N = 6$ none of the methods converged when $\omega =
        0.1$\footnote{The RHF-method would probably have converged if we had
        started tuning its mixing parameter, but as we have focused on CCD we did
        not think it relevant to include these results.}. By increasing the
        frequency we restored the convergence. For larger number of shells we
        had to start increasing the value of $\theta \to 0.9$ in order to get
        convergence within the threshold.

        CCD with HO basis experiences the same convergence problems also for,
        $N=12$ particles and for $N=20$ particles.

    %\subsection{Sparse implementation}
    % Did you want to write something about memory use, Oyvind? Memory problems?

    \section{Summary Remarks}

        Our main product from this is a product that can do both computations with
        restricted Hartree-Fock (RHF) self-consistent field and coupled cluster
        doubles (CCD). We have tried
        to make a package in python that should be easy enough for anyone to test.
        Moreover we hav demonstrated that one can speed up the computation time with
        a relatively simple algebraic transformation and finding the ``intermediates''.
        The computational speed was further reduced using sparse matrices. It should
        be noted, as a warning, that there are relatively few systems adequately sparese
        for this to work. We added upon the program to increase it further with
        paralellization and optimal use of numerical functions.

        In order to perform the computations we have constructed a harmonic oscillator
        basis. We saw how the CCD method was outperformed RHF in most cases, and often
        exhibited convergence problems. The remedy for these shortcomings is to use
        the Hartree-Fock basis which is provided after successful convergence of
        SCF iterations. The results are improved markedly and classification of coupled
        cluster methods as a post-Harte-Fock method is justified.

        The coupled cluster method provides surprisingly good results even
        though one only includes double esctiations. The reason for this is that
        particle-particle interactions is the amongt the most important thing that
        happens in a ferminon system, second only to no interaction at all, i.e
        what particles do on their own. The way to improve the computations is to
        include other types of excitations like singles (CCSD) and triples (CCSDT),
        add more basis functions, and/or supplement the coupled cluster method with
        another method like a perturbative one. All of this will, needless to say,
        increase the compute time, which must be balanced against the accuracy needed
        for the situation at hand.

\vfill
\newpage
\appendix
\section{The normal ordered Hamiltonian}
    When constructing the normal ordered Hamiltonian we use Wick's theorem to
    write the one-body, $h$, and the two-body, $u$, operators onto a normal
    ordered form. Specifically we define the normal ordered form in terms of the
    \emph{Fermi vacuum}\footnote{Fermi vacuum defines the reference state, i.e.,
    $\kslat$, as the vacuum.}. That is, an operator on normal ordered form
    destroys the reference Slater determinant.

    We start by writing the one-body operator, $h$, to its normal-ordered form.

@@

        h &= \sum_{pq}h^p_q\acr{p}\ade{q}
        = \sum_{pq}h^p_q\para{

@@

        &= \sum_{pq}h^p_q\{

@@

        + \sum_{pq}h^p_q\delta_{p \in i}\delta_{pq}

@@

        &= h_N + \sum_{i}h^i_i,

@@

    endeavor. For brevity we will only write out the operator strings and only
    keep the non-zero contributions.

@@

        \frac{1}{4}\sum_{pqrs}\bra{pq}\ket{rs}

@@

        - \frac{1}{4}\sum_{iqr}\bra{iq}\ket{ri}\{\acr{q}\ade{r}\}

@@

        + \frac{1}{4}\sum_{iqs}\bra{iq}\ket{is}\{\acr{q}\ade{s}\}
        + \frac{1}{4}\sum_{pir}\bra{pi}\ket{ri}\{\acr{p}\ade{r}\}

@@

        - \frac{1}{4}\sum_{pis}\bra{pi}\ket{is}\{\acr{p}\ade{s}\}
        - \frac{1}{4}\sum_{ij}\bra{ij}\ket{ji}

@@

        + \frac{1}{4}\sum_{ij}\bra{ij}\ket{ij}.

@@

        \bra{pq}\ket{rs}
        = - \bra{pq}\ket{sr}
        = - \bra{qp}\ket{rs}
        = \bra{qp}\ket{sr},

@@

        W_N + \sum_{pir}\bra{pi}\ket{ri}\{\acr{p}\ade{r}\}
        + \half\sum_{ij}\bra{ij}\ket{ij},

@@

        W_N = \frac{1}{4}\sum_{pqrs}
        \bra{pq}\ket{rs}\{\acr{p}\acr{q}\ade{s}\ade{r}\}.

@@

    constants in both the one-body and the two-body operator in total constitues
    the reference energy.

@@

        = \sum_{i}h_i^i + \frac{1}{2}\sum_{ij}\bra{ij}\ket{ij}.

@@

        \sum_{pq}h_{q}^{p}\{\acr{p}\ade{q}\}
        + \sum_{pqi}\bra{pi}\ket{qi}\{\acr{p}\ade{q}\}

@@

        &= \sum_{pq}f_{q}^{p}\{\acr{p}\ade{q}\},

@@

        h_q^p + \sum_{i}\bra{pi}\ket{qi}.

@@

\section{Energy equation}
    \label{app:energy_equation}
    Here we show how to compute the coupled cluster doubles energy from
    \autoref{eq:ccd_energy}. The first two terms come from the reference energy
    shown in \autoref{eq:reference_energy}. The last term we get by using Wick's
    theorem.
    \begin{align}
        \bslat (H_N T_2)_c\kslat
        &= \bslat (W_N T_2)_c\kslat,
    \end{align}
    as the normal ordered Fock operator can at most relax a single state, i.e.,
    \begin{align}
        \bslat (F_N T_2)_c\kslat = 0.
    \end{align}
    We are thus left with computing the contraction of the doubles cluster
    operator and the normal ordered two-body operator. Note that we must keep
    only the fully contracted terms.
    \begin{align}
        \bslat (W_N T_2)_c\kslat
        &=
        \frac{1}{16} u^{pq}_{rs} t^{ab}_{ij}
        \bslat
        A
        \kslat,
        \label{eq:third_term_ccd_energy}
    \end{align}
    where $A$ is the create and destruction operator strings. By using
    generalized Wick's theorem we get
    \begin{align}
        A &=
        \{\acr{p}\acr{q}\ade{s}\ade{r}\}
        \{\acr{a}\acr{b}\ade{j}\ade{i}\}
        \nonumber \\
        &=
        \{
            \wick{
                    \c4 a_p^{\dagger} \c3 a_q^{\dagger}
                    \c2 a_s \c1 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        +
        \{
            \wick{
                    \c4 a_p^{\dagger} \c3 a_q^{\dagger}
                    \c1 a_s \c2 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        \nonumber \\
        &\qquad
        +
        \{
            \wick{
                    \c3 a_p^{\dagger} \c4 a_q^{\dagger}
                    \c2 a_s \c1 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        +
        \{
            \wick{
                    \c3 a_p^{\dagger} \c4 a_q^{\dagger}
                    \c1 a_s \c2 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        \nonumber \\
        &=
        \delta_{pi}\delta_{qj}\delta_{sb}\delta_{ra}
        - \delta_{pi}\delta_{qj}\delta_{sa}\delta_{rb}
        \nonumber \\
        &\qquad
        - \delta_{pj}\delta_{qi}\delta_{sb}\delta_{ra}
        + \delta_{pj}\delta_{qi}\delta_{sa}\delta_{rb}.
    \end{align}
    Inserted back into \autoref{eq:third_term_ccd_energy} and summing the
    indices in the Kronecker-Delta's we get
    \begin{align}
        \bslat (W_N T_2)_c\kslat
        &=
        \frac{1}{16}\para{
            u^{ij}_{ab} - u^{ij}_{ba} - u^{ji}_{ab} + u^{ji}_{ba}
        }
        t^{ab}_{ij}
        \nonumber \\
        &=
        \frac{1}{4}u^{ij}_{ab}t^{ab}_{ij},
    \end{align}
    where we have used the antisymmetry of the two-body matrix elements, thus
    completing what we wanted to show.

@@

    In this section we have provided a few sample computations of how one would
    evaluate the amplitude equations using wicks theorem. Starting with the simplest
    term including only the normal-ordered Hamiltonian,

    \begin{equation}
        \begin{aligned}
            \bra{\Phi_{ij}^{ab}} &(F_N + V_N)\ket{\Phi_0} \\
             =& \sum_{pq}f_{pq}\bra{\Phi_0}\{a_i^\dagger a_j^\dagger a_ba_a\}\{a^\dagger_pa_q\}\ket{\Phi_0} \\
             &+\frac{1}{4}\sum_{pqrs}\bra{pq}\ket{rs}
             \bra{\Phi_0}\{a_i^\dagger a_j^\dagger a_ba_a\}\{a^\dagger_pa^\dagger_qa_sa_r\}\ket{\Phi_0}.
        \end{aligned}
        \end{equation}
        The one-electron component does not have any full contractions, while the
        two-electron component produces one contributing integral,
        \begin{equation}
        \begin{aligned}
            \bra{\Phi_{ij}^{ab}} &(V_N)\ket{\Phi_0} \\
                =& \frac{1}{4}\sum_{pqrs}\bra{pq}\ket{rs}
             \bra{\Phi_0}\{a_i^\dagger a_j^\dagger a_ba_a\}\{a^\dagger_pa^\dagger_qa_sa_r\}\ket{\Phi_0} \\
                 =& \frac{1}{4} \sum_{pqrs} \bra{pq}\ket{rs} \\ &\times\Big(
                     \{ \wick{\c1 a^\dagger_i \c2 a^\dagger_j \c3 a_b \c4 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                     +
                     \{ \wick{\c1 a^\dagger_i \c2 a^\dagger_j \c4 a_b \c3 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\} \\
                   &\ +
                      \{ \wick{\c2 a^\dagger_i \c1 a^\dagger_j \c3 a_b \c4 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                     +
                     \{ \wick{\c2 a^\dagger_i \c1 a^\dagger_j \c4 a_b \c3 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                 \Big) \\
                 =&  \frac{1}{4} \sum_{pqrs} \bra{pq}\ket{rs}
                 ( \delta_{ap}\delta_{bq}\delta_{js}\delta_{ir}
                  -\delta_{aq}\delta_{bp}\delta_{js}\delta_{ir} \\
                 &\ -\delta_{ap}\delta_{bq}\delta_{jr}\delta_{is}
                  +\delta_{aq}\delta_{bp}\delta_{jr}\delta_{is}) \\
                  =& \bra{ab}\ket{ij} = u^{ab}_{ij}.
        \end{aligned}
        \end{equation}

        Next we would like to evaluate $\bra{\Phi_{ij}^{ab}}(F_N + V_N)T_2\ket{\Phi_0}$. Starting with the
        term involving the Fock operator $F_N$,
        \begin{equation}
        \begin{aligned}
            \bra{\Phi_{ij}^{ab}}&F_NT_2\ket{\Phi_0} \\
                =& \frac{1}{4}\sum_{pq}\sum_{klcd}f_{pq}t_{kl}^{cd} \bra{\Phi_0}
                    \{a_i^\dagger a_j^\dagger a_b a_a \}(\{a^\dagger_pa_q \}\{a^\dagger_c a^\dagger_d a_l a_k \})_c \ket{\Phi_0} \\

>>>>

    Regardless of how sophisticated the method used, the many-body problem
    remains quite intractable. The true wavefunction of any system will be a
    linear combination of all possible slater-type orbitals one can model. This
    means that in order to represent a system correctly, one would need a basis
    set of infinite size. This is an obvious problem when we solve problems
    numerically, because we obviously need to limit the size of the basis we use
    because we have limitations in memeory and computating power. Luckily, great
    minds have bestowed a wide selection of very good numerical methods upon the
    world, which the coupled cluster is a part of.

    First, we provide a thorough theoretical outline of how we model a quantum
    dot, the mathematical formalism behind many-body problems and outline how
    the computational methods we employ both turn into iterative schemes that
    are easy to represent numerically. Second, we provide a brief overview of
    the program we have constructed which is in the form an installable Python
    package.  Third, we have employed the methods on a large parameter space
    presenting results for systems consisting of $N= \{2, 6, 12, 20\}$
    particles. Fourth, we discuss the most interesting aspects of the results
    and try to shed some light over the shortcomings as well as the beneficial
    features of the different metods. Lastly, we end on some summary remarks.

@@

        The theoretical approximation of a quantum dot is impossible to make
        perfectly accurate, yet simple. There are some important effects that
        should be included such as the potential arising as a result of the
        confinement, and the interaction between particles. We therefore model
        our system with an electron-electron interaction, given by the two-body
        operator in \autoref{eq:two_body_operator}, and with it we approximate
        the confining potential by the harmonic oscillator potential, given by
        the second term in the one-body Hamiltonian in
        \autoref{eq:one_body_hamiltonian}. The first term in the one-body
        Hamiltonian is the kinetic energy of particle in question, essential in
        any system.

@@

    \subsection{Implementation details}
        Both in \texttt{CoupledClusterDoublesSparse} and
        \texttt{CoupledClusterDoublesOptimized} we make use of the
        intermediates in equations \ref{eq:intermediate1} to
        \ref{eq:intermediate4}.  In the optimized class special care was taken
        to ensure that every contraction was computed in an optimal way, with
        the minimization of memory use and CPU time as the overall goal.
        Moreover, several methods in NumPy for doing linear algebra was tested,
        including but not limited to \texttt{numpy.einsum},
        \texttt{numpy.tensordot}, and reshaping datastructures to
        two dimensions and using \texttt{numpy.matmul}.
        Additionally, we tried to make use of the Python package
        \texttt{dask} which allows for arrays that are larger than
        the computer's memory buffer as well as automated concurrency. We found
        that the dask arrays was a bad fit for our code base, but we ended up
        using \texttt{numba} for paralellization of some
        iterative loops as well as just-in-time compilation of some functions.

    \subsection{Scaling of the coupled cluster doubles method}
        When talking about the scaling of the coupled cluster method there are
        two aspects to consider.
        \begin{enumerate}
            \item The number of FLOPS to compute $g(f, u, t)$.
            \item The number of bytes needed to store the arrays of matrix
                elements.
        \end{enumerate}
        For the CCD-method the naïve solution using the form of $g(f, u, t)$
        without the intermediates results in $\mathcal{O}(M^4N^4)$ FLOPS each
        iteration. Using the form of the intermediates we have set up this
        reduces to $\mathcal{O}(M^4N^2)$, i.e., two order of magnitude less. The
        effect of this is \emph{dramatic} and will be shown in the results
        section in due time.

@@

    \subsection{Validity of results}
        The analytical computed energy for $\omega=1.0$ a.u.  and $N=2$ is $3$,
        from Taut\cite{taut1994two}. From our tables we see that RHF is the
        least successful in reaching this value, CCD using the HF basis is best,
        while CCD with the HO basis is generally somewhere between the two. We
        see that we get even closer to the analytical ``truth'' by increasing
        the number of shells.

        As predicted in the previous subsection we see that the CCD-method using
        the HO basis gradually deteroriates compared with RHF and CCD using the
        HF basis as the number of particles increases. In fact, it quickly
        reaches a point where we get trouble with convergence due to the
        unstable states with a large number of particles packed together in a
        too weak trap.

        When the states become too unstable we have to increase the mixing
        parameter $\theta \to 1$ in \autoref{eq:mixing}. This will sometimes
        make the method converge, but it can be at the cost of a wrong value of
        the energy as we might get trapped in unstable configurations which with
        the help of mixing will converge for our criteria.

        The rest of the results have been compared with the master thesis of M.
        P.  Lohne\cite{lohne2010coupled} for up to $R = 10$ shells. Lohne gets
        two sets of energies, one set from RHF and one set from CCSD code using
        harmonic oscillator basis functions. Our CCD code with Hartree-Fock
        basis will for some configurations\footnote{By configurations we mean
        frequency $\omega$ and number of particles $N$.} beat CCSD with plain
        harmonic oscillator basis. For $R = 12$ shells we can compare with the
        results from M. P. Lohne et al.\cite{lohne2011ab}, but in this article
        an \emph{effective interaction} for the Hamiltonian with a CCSD code
        using the HF basis has been used. These results are therefore
        significantly better than ours, but we can get a ``ball-park'' idea to
        benchmark against.

    \subsection{Running time}
        Looking at the running times shown in \autoref{fig:running_time_2} and
        \autoref{fig:running_time_6} we can see how the naïve implementation
        blows up very fast. As both the sparse and the optimized version uses
        intermediate calculations these running times scales on the order of
        $\mathcal{O}(M^4N^2)$ whereas the naïve implementation goes as
        $\mathcal{O}(M^4N^4)$. The exponential difference in scaling is readily
        apparent from these plots. Another important factor is that the naïve
        implementation uses \texttt{np.einsum} to calculate the tensor
        contractions whereas the sparse and optimized version uses
        \texttt{sparse.tensordot} and \texttt{np.matmul} respectively. The two
        latter function calls uses BLAS's implementation of the matrix dot which
        is way faster than Numpys \texttt{np.einsum}.

    \subsection{Convergence trouble}
        For a large number of particles and a low frequency the CCD-method get
        trouble with convergence. This happens as the confining potential $v
        \propto \omega^2$ will not be able to confine the particles when the
        interaction gets too strong. The CCD-method will in particular get a
        hard time keeping the particles together as the excitation operator $t$
        only excites pairs of particles leading to a too strong increase in the
        interaction when the particles increase their energy level. This effect
        can potentially be somewhat alleviated by including the singles
        excitation, e.g., using the CCSD-method.

        We can see this behaviour in our results. For $N = 2$ we get convergence
        for all methods for $\omega = 0.1$ shown in \autoref{tab:N2}. By
        increasing to $N = 6$ particles we immediately see how CCD with the HO
        basis will get in trouble for small values of $\omega$. Even by
        increasing the strength of the confinment this method does not always
        converge.

        For $N = 6$ none of the methods converged when $\omega =
        0.1$\footnote{The RHF-method would probably have converged if we had
        started tuning its mixing parameter, but as we have focused on CCD we did
        not think it relevant to include these results.}. By increasing the
        frequency we restored the convergence. For larger number of shells we
        had to start increasing the value of $\theta \to 1$ in order to get
        convergence within the threshold.

        The CCD-method with the HO basis experiences the same convergence
        problems also for, $N=12$ particles and for $N=20$ particles.

    \subsection{Sparse implementation}
        From \autoref{fig:running_time_6_zoom} we can see that the sparse
        implementation is close to the optimized version of the CCD-method. This
        is a sobering idea as the sparse implementation is by no means optimized
        properly and more resembles the naïve implementation. This brings faith
        in the fact that a more optimized version of the sparse implementation
        using block matrices will be able to achieve much better performance.
        This is however something that will be experimented with in the future.

\section{Summary Remarks}
    Our main product from this project is a code that can do both computations
    using the restricted Hartree-Fock (RHF) self-consistent field and coupled
    cluster doubles (CCD). We have tried to make a package in Python that should
    be easy enough for anyone to test.  Moreover we have demonstrated that one
    can speed up the computation time with a relatively simple algebraic
    transformation and finding the ``intermediates''. The computational speed
    was further reduced using sparse matrices. It should be noted, as a warning,
    that not all systems are sparse enough for a sparse implementation to make
    sense.  We added upon the program to increase it further with
    paralellization and optimal use of numerical functions.

    In order to perform the computations we have constructed a harmonic
    oscillator basis. We saw how the CCD-method using this basis was
    outperformed by RHF in most cases, and often exhibited convergence problems.
    The remedy for these shortcomings is to use the Hartree-Fock basis which is
    provided after successful convergence of SCF iterations. The results are
    improved markedly and the classification of coupled cluster methods as a
    post-Harte-Fock method is justified.

    The coupled cluster method provides surprisingly good results even though
    one only includes double exitations. The reason for this is that
    particle-particle interactions is among the most important thing that
    happens in a fermionic system, second only to no interaction at all, i.e
    what particles do on their own. The way to improve the computations is to
    include other types of excitations like singles (CCSD) and triples (CCSDT),
    add more basis functions, and/or supplement the coupled cluster method with
    another method like perturbation theory. All of this will, needless to say,
    increase the compute time, which must be balanced against the accuracy
    needed for the situation at hand.

\vfill
\newpage
\appendix
\section{The normal ordered Hamiltonian}
    When constructing the normal ordered Hamiltonian we use Wick's theorem to
    write $h$ and $u$ onto a normal ordered form. Specifically we define the
    normal ordered form in terms of the \emph{Fermi vacuum}\footnote{Fermi
    vacuum defines the reference state, i.e., $\kslat$, as the vacuum.}. That
    is, an operator on normal ordered form destroys the reference Slater
    determinant.

    We start by writing the one-body operator $h$, to its normal-ordered form.

@@

        h &= h^p_q\acr{p}\ade{q}
        = h^p_q\para{

@@

        &= h^p_q\{

@@

        + h^p_q\delta_{p \in i}\delta_{pq}

@@

        &= h_N + h^i_i,

@@

    endeavor. For brevity we will only write out the operator strings.

@@

        \frac{1}{4}u^{pq}_{rs}

@@

        - \frac{1}{4}u^{iq}_{ri}\{\acr{q}\ade{r}\}

@@

        + \frac{1}{4}u^{iq}_{is}\{\acr{q}\ade{s}\}
        + \frac{1}{4}u^{pi}_{ri}\{\acr{p}\ade{r}\}

@@

        - \frac{1}{4}u^{pi}_{is}\{\acr{p}\ade{s}\}
        - \frac{1}{4}u^{ij}_{ji}

@@

        + \frac{1}{4}u^{ij}_{ij}.

@@

        u^{pq}_{rs}
        = - u^{pq}_{sr}
        = - u^{qp}_{rs}
        = u^{qp}_{sr},

@@

        W_N + u^{pi}_{ri}\{\acr{p}\ade{r}\}
        + \half u^{ij}_{ij},

@@

        W_N = \frac{1}{4}
        u^{pq}_{rs}\{\acr{p}\acr{q}\ade{s}\ade{r}\}.

@@

    constants in both the one-body and the two-body operator in total
    constitutes the reference energy.

@@

        = h_i^i + \frac{1}{2}u^{ij}_{ij}.

@@

        h_{q}^{p}\{\acr{p}\ade{q}\}
        + u^{pi}_{qi}\{\acr{p}\ade{q}\}

@@

        &= f_{q}^{p}\{\acr{p}\ade{q}\},

@@

        h_q^p + u^{pi}_{qi}.

@@

\section{Energy equation}
    \label{app:energy_equation}
    Here we show how to compute the coupled cluster doubles energy from
    \autoref{eq:ccd_energy}. The first two terms come from the reference energy
    shown in \autoref{eq:reference_energy}. The last term we get by using Wick's
    theorem.
    \begin{align}
        \bslat (H_N T_2)_c\kslat
        &= \bslat (W_N T_2)_c\kslat,
    \end{align}
    as the normal ordered Fock operator can at most relax a single state, i.e.,
    \begin{align}
        \bslat (F_N T_2)_c\kslat = 0.
    \end{align}
    We are thus left with computing the contraction of the doubles cluster
    operator and the normal ordered two-body operator. Note that we must keep
    only the fully contracted terms.
    \begin{align}
        \bslat (W_N T_2)_c\kslat
        &=
        \frac{1}{16} u^{pq}_{rs} t^{ab}_{ij}
        \bslat
        A
        \kslat,
        \label{eq:third_term_ccd_energy}
    \end{align}
    where $A$ is the creation and destruction operator strings. By using the
    generalized Wick's theorem we get
    \begin{align}
        A &=
        \{\acr{p}\acr{q}\ade{s}\ade{r}\}
        \{\acr{a}\acr{b}\ade{j}\ade{i}\}
        \nonumber \\
        &=
        \{
            \wick{
                    \c4 a_p^{\dagger} \c3 a_q^{\dagger}
                    \c2 a_s \c1 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        +
        \{
            \wick{
                    \c4 a_p^{\dagger} \c3 a_q^{\dagger}
                    \c1 a_s \c2 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        \nonumber \\
        &\qquad
        +
        \{
            \wick{
                    \c3 a_p^{\dagger} \c4 a_q^{\dagger}
                    \c2 a_s \c1 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        +
        \{
            \wick{
                    \c3 a_p^{\dagger} \c4 a_q^{\dagger}
                    \c1 a_s \c2 a_r
                    \c1 a_a^{\dagger} \c2 a_b^{\dagger}
                    \c3 a_j \c4 a_i
            }
        \}
        \nonumber \\
        &=
        \delta_{pi}\delta_{qj}\delta_{sb}\delta_{ra}
        - \delta_{pi}\delta_{qj}\delta_{sa}\delta_{rb}
        \nonumber \\
        &\qquad
        - \delta_{pj}\delta_{qi}\delta_{sb}\delta_{ra}
        + \delta_{pj}\delta_{qi}\delta_{sa}\delta_{rb}.
    \end{align}
    Inserted back into \autoref{eq:third_term_ccd_energy} and summing the
    indices in the Kronecker-Delta's we get
    \begin{align}
        \bslat (W_N T_2)_c\kslat
        &=
        \frac{1}{16}\para{
            u^{ij}_{ab} - u^{ij}_{ba} - u^{ji}_{ab} + u^{ji}_{ba}
        }
        t^{ab}_{ij}
        \nonumber \\
        &=
        \frac{1}{4}u^{ij}_{ab}t^{ab}_{ij},
    \end{align}
    where we have used the antisymmetry of the two-body matrix elements, thus
    completing what we wanted to show.

@@

    In this section we have provided a few sample computations of how one would
    evaluate the amplitude equations using Wick's theorem. Starting with the
    simplest term including only the normal-ordered Hamiltonian,

    \begin{equation}
        \begin{aligned}
            \bra{\Phi_{ij}^{ab}} &(F_N + V_N)\ket{\Phi_0}
             = f^p_q\bra{\Phi_0}\{a_i^\dagger a_j^\dagger a_ba_a\}
             \{a^\dagger_pa_q\}\ket{\Phi_0} \\
             &\qquad
             + \frac{1}{4}u^{pq}_{rs}
             \bra{\Phi_0}\{a_i^\dagger a_j^\dagger a_ba_a\}
             \{a^\dagger_pa^\dagger_qa_sa_r\}\ket{\Phi_0}.
        \end{aligned}
    \end{equation}
    The one-electron component does not have any full contractions, while the
    two-electron component produces one contributing integral,
    \begin{equation}
        \begin{aligned}
            \bra{\Phi_{ij}^{ab}} &(V_N)\ket{\Phi_0}
            = \frac{1}{4}u^{pq}_{rs}
            \bra{\Phi_0}\{a_i^\dagger a_j^\dagger a_ba_a\}
            \{a^\dagger_pa^\dagger_qa_sa_r\}\ket{\Phi_0} \\
                 &= \frac{1}{4} u^{pq}_{rs} \\
                 &\times
                 \Biggl(
                     \{ \wick{\c1 a^\dagger_i \c2 a^\dagger_j \c3 a_b \c4 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                     +
                     \{ \wick{\c1 a^\dagger_i \c2 a^\dagger_j \c4 a_b \c3 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                     \\
                     &+
                      \{ \wick{\c2 a^\dagger_i \c1 a^\dagger_j \c3 a_b \c4 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                     +
                     \{ \wick{\c2 a^\dagger_i \c1 a^\dagger_j \c4 a_b \c3 a_a
                     \c4 a^\dagger_p \c3 a^\dagger_q \c2 a_s \c1 a_r}\}
                 \Biggr) \\
                 &=  \frac{1}{4} u^{pq}_{rs}
                 \biggl( \delta_{ap}\delta_{bq}\delta_{js}\delta_{ir}
                 -\delta_{aq}\delta_{bp}\delta_{js}\delta_{ir} \\
                 & -\delta_{ap}\delta_{bq}\delta_{jr}\delta_{is}
                 + \delta_{aq}\delta_{bp}\delta_{jr}\delta_{is}\biggr) \\
                &= u^{ab}_{ij}.
        \end{aligned}
        \end{equation}

        \begin{widetext}
            Next we would like to evaluate the normal ordered Hamiltonian connected
            with a single doubles cluster operator. Starting with the term involving
            the Fock operator $F_N$,
                \bra{\Phi_{ij}^{ab}}(F_NT_2)_c\ket{\Phi_0}
                    &= \frac{1}{4}f^p_qt_{kl}^{cd} \bra{\Phi_0}
                    \{a_i^\dagger a_j^\dagger a_b a_a \}
                    \para{
                        \{a^\dagger_pa_q \}\{a^\dagger_c a^\dagger_d a_l a_k \}
                    }_c \ket{\Phi_0}
                    &= \frac{1}{4}f^p_q t_{kl}^{cd}
                    \Biggl(
                        \{
                            \wick{
                                \c1 a^\dagger_i \c2 a^\dagger_j \c3 a_b
                                \c4 a_a \c5 a^\dagger_p \c2 a_q \c3 a^\dagger_c
                                \c4 a^\dagger_d \c1 a_l \c5 a_k
                            \wick{
